Static Analysis by Path Focusing

### 2011-06-13 report/report.tex
report

Static analysis aims at automatically computing properties on programs, such as
the possible values of their variables during execution. This allows to show for
instance that a program will not overflow, will not divide by zero, and to compute
loop invariants\dots
The main applications of static analysis are compilation-time optimisation, and
the proof of safety properties in critical systems, such as avionics.

@@

In this report we make the following contributions:
\begin{itemize}
\item We present a new technique, referred as \emph{Path Focusing}, that takes
benefits of the recent advances in the field of SMT-solving to improve precision
of static analysis by abstract interpretation. We also present an adaptation of
this technique to compute disjunctive invariants.
\item We present an implementation of this technique into a small analyser, as
well as the implementation of another technique proposed in \cite{GopanR06},
such that we can compare precision and cost of these two methods.
\end{itemize}

@@

Static analysis is aimed at discovering some properties about programs. In this
example, one would compute the set of possible values for the variable $x$
during execution.
The graph at the right side represents the program. The nodes of the graph are
called \emph{program points}. We can compute for instance the set $Y_1$ of
values for $x$ at point $p_1$. This set depends on the sets $Y_0$ and $Y_2$,
since there are two edges arriving at $p_1$: the first one comes from $p_0$, and
the second one from $p_2$.

@@

However, most of the time, the solution of this fixpoint equation must be
computed in a complex domain: in program analysis, this could be the state space
of the program, i.e the set of all possible states of the program. This
computation quickly becomes too costly or doesn't terminate.

@@

Termination of the fixpoint computation has to be guaranteed. This termination
depends on the properties of the abstract domain:
this one should be finite, of finite height,
or more generally satisfy the \emph{ascending chain condition},
meaning that there doesn't exist
any sequence $(x_i)_{i \geq 0}$ of elements of the abstract domain $A$ such that
$\forall i \geq 0,\ x_i <_A x_{i+1}$. Indeed, otherwise, the least fixpoint
computation could run indefinitely, since the ascending sequence of elements of
$A$ is infinite, and the fixpoint in never reached.

Many of the domains that are used for program analysis, including those of intervals and convex polyhedra, don't satisfy this ascending chain condition. To ensure convergence in this case, another approximation
is performed: a new operator is defined, called \emph{widening operator}, that
extrapolates the limit of a sequence of abstract values
\cite{CC77,CousotCousot92-4}. This \emph{widening
operator} is usually noted $\widening: A \times A \rightarrow A$, and satisfies
the following properties:

@@

Let $x_1, x_2, \dots ,x_n$ be the numerical variables of a program (assume they
all are in $\Q$). A state of
the program is then a point $\overrightarrow{x} \in \Q^n$.

@@

There have been various
work proposing a definition or a refinement of the widening operator in LRA
\cite{CH78,Hal79,HPR97,BlanchetCousotEtAl_PLDI03}, in order to fight
against loss of precision:
\begin{itemize}
\item \emph{Delayed widening}: instead of applying widening at the first
iteration of the analysis, one apply it after a number $n$ of iterations. During
the $n-1$ first iterations, only a convex hull is computed. In some cases,
delaying the widening operation is more precise. The algorithm still terminates,
since widening is applied within a finite time after $n$ iterations. In most of
the cases, $n$ can be chosen equals to $2$.
\item \emph{Widening with threshold (or ``up to'')}: In some cases, instead of
brutally applying the widening operator, a fixed set of linear inequalities $T$
can be chosen, and the widening operator $\widening_T$ is defined as follows:
$P \widening_T Q$ is the intersection of $P \widening Q$ with all the inequalities in
$T$ satisfied by both $P$ and $Q$.

@@

To each state $p \in P$ of the control flow graph, we associate an abstract
value $X_p \in D$, $D$ being in our case the domain of convex polyhedra over
$\Q^n$. Since the exact operation $\tau$ may not be expressed in the abstract
domain, we abstract it into $\tau^\#$ such that $\forall X \in D, \tau(X)
\subseteq \tau^\#(X)$.

@@

After a finite number of steps, the computation has reached the fixpoint and the
value $X_p$ at the end give the different linear relations between the
different numerical variables of the program at the control point $p$.

\subsection{Precision of the analysis}
In order to guarantee the termination of the computation, there is no need to
apply the widening operator at each control point $p \in P$. It is sufficient to
apply it on a subset of $P$, called $P_w$, such that removing the nodes of $P_w$
disconnects every cycles. For instance, $P_w$ can be chosen as the set of the
heads of loops.

\subsubsection{Iteration strategies}
There exist several iteration strategies for computing the fixpoint: whatever
order we choose for updating the different $X_p, p\in P$, the result at the
end of the analysis is correct. Yet, the precision of the result and the time
before convergence could be very different. 
There have been some work to find efficient iteration strategies
\cite{Bou92}, such as first stabilising innermost loops, or stabilising strongly
connected components of the graph.

@@

	If the problem has a solution,  the SMT-solver will be in position to show
	which path in the control flow graph is selected, by giving a model: an
	assignments of the Boolean variables so that the formula is true.
	Among these Boolean variables, some of them are associated to a transition
	in the control flow graph:
	\begin{itemize}
	\item if the model has set to \emph{true} the Boolean variable $b_e$ associated
	to the transition $e$, then the focus path goes through this transition.
	\item in the other case, the focus path doesn't go through this transition.
	\end{itemize}

@@

\begin{itemize}
\item Step 0: Initially, $X_0 = \top, X_1 = \perp, X_7 = \perp$, and $A =
\{p_0\}$.
\item Step 1: We start in $p_0$. Is there a path starting in $p_0$, that arrives
in a state in $P_R$ and makes its abstract value grow? The SMT-solver answers
\emph{yes}, and gives the path $p_0 \rightarrow p_1$. We update $X_1$:
it is the polyhedron $x=y=0$. We update $A = \{p_1\}$. 
We do another SMT query: is there another path
starting in $p_0$ that makes an abstract value grow? The answer is no.
\item Step 2: We now work on $p_1$. Is there a path starting in $p_1$ that
arrives in a successor of $p_1$, and makes its abstract value grow? Answer is
yes, and the model given by the SMT-solver is the path $p_1 \rightarrow p_2
\rightarrow p_4 \rightarrow p_5 \rightarrow p_6 \rightarrow p_1$. Actually, this
is the first phase of the loop. 
The image of $x=y=0$ by this transition is $x=y=1$. The new $X_1$ is then $x=y
\wedge 0\leq x \leq 1$. The path is a self loop, but is seen for the first time,
so we do not apply widening yet. So, we redo an SMT-query, that gives us exactly
the same path. Then, we can do a widening/narrowing sequence. 
After widening, we have $x=y \wedge x \geq 0$, and narrowing gives $X_1 = (x=y)
\wedge (0 \leq x \leq 51)$.
\item Step 3: We still are in $p_1$: we ask for another focus path, and the
SMT-solver finds the path $p_1 \rightarrow p_3
\rightarrow p_4 \rightarrow p_5 \rightarrow p_6 \rightarrow p_1$. This is the
second phase of the loop, which is possible since $(x,y)=(51,51) \in X_1$.
Again, this is a self-loop, but we do not widen yet. The
image of $X_1$ by the transition is the single point $(52,50)$. $X_1$ is then
the convex hull of this point with the old $X_1$. A new SMT query gives the same
path, and we can now apply widening/narrowing. We obtain
$X_1 = (x \leq y) \wedge (102 - x - y \geq 0)$ after widening, and narrowing
gives $X_1 = (x \leq y) \wedge (102 - x - y \geq 0) \wedge y \geq 0$.
We can see that $X_1$ is the smallest polyhedron containing the possible values
for $(x,y)$ (see Figure \ref{result}.b).
\item Step 4: Again, we ask the SMT-solver for a new focus path, and it returns
$p_1 \rightarrow p_3 \rightarrow p_4 \rightarrow p_7$. The image of the
polyhedra $X_1$ after this path transformation is the single point $y=-1 \wedge
x = 102$. We can update $X_7 = {(102,-1)}$. Path focusing technique gives us the
precise result of this program. After that, the SMT-solver can't find any new
path, so the analysis terminates.
\end{itemize}

@@

The new algorithm we propose is described in Algorithm \ref{gulwani2}.
A difference is that here, we don't know the number of paths. Then, the mapping
function $\sigma$ has to be defined on $\{1,..,m\} \times \N$. We could also
easily compute this number of paths when computing the $\rho$ formula.

@@

\paragraph{Path Focusing} 
The analysis of the loop starts with 
$p_1: x=i=r=0$. Then, the SMT-solver discovers the path $p_1 \rightarrow p_2
\rightarrow p_3 \rightarrow p_5 \rightarrow p_7 \rightarrow p_1$. We do a
widening/narrowing sequence and we find 
$ 0 \leq i \leq 50 \wedge r=0 \wedge -1 \leq x \leq 0$ at $p_1$. The SMT-solver
gives another path: 
$p_1 \rightarrow p_2
\rightarrow p_4 \rightarrow p_5 \rightarrow p_7 \rightarrow p_1$. After
analysing this path, we obtain 
$ 0 \leq i \leq 100 \wedge r=0 \wedge x=0$. Analysis terminates since the
SMT-solver gives $unsat$ answers. Here, We never focused on paths going through
point $p_6$, so our analysis gives us the invariant $x=0$ at the end of the
program. Indeed, there is no feasible paths going through $p_6$, but
\emph{Lookahead Widening} don't know it because of the convex hull at point
$p_5$.

@@

LLVM internal representation can be applied transformation passes that modifies
parts of it. These passes are mostly used for optimisation purpose. In our
analyser, we apply some already existing passes before running our 
abstract interpretation pass. Among the passes we run, we could cite:
\begin{itemize}
\item the \emph{PromoteMemoryToRegister} Pass, that transforms the internal
representation by promoting the memory variables to Static Single Assignment
(SSA) registers. This pass transform the control flow graph such that all the
variables are in SSA form, and removes most of the \emph{Load/Store} operations
into memory. It is useful since our analyser doesn't have a memory model, and
then loses lots of precision in the case of \emph{Load/Store} instructions.
\item the \emph{LoopSimplify} and \emph{LoopInfo} Passes, that transform all loops into a canonical
format, and compute some information related to loops (for instance, the set of
BasicBlocks that are parts of the loop, etc.).
\item the \emph{LowerSwitch} Pass, that transform a switch instruction into its
corresponding \emph{if-then-else} instructions. After running this pass, there
is no more \emph{Switch} instruction in the control flow graph, so the
implementation of the static analysis is easier.
\end{itemize}

@@

This is the case for the distinction between \emph{signed} and \emph{unsigned}
integers, which is nonexistent in the representation, since the compiler doesn't
need this information to generate code. Yet, for static analysis, 
this distinction is required to improve precision:
for instance, if $n$ is unsigned, the constraint $n \geq 0$ could be added in
	our abstract value at the beginning of the analysis.
	This lack of precision may be limited by using disjunctive invariants: one
	could attach two polyhedra to the control point: the first one with the
	constraint $n \geq 0$, and the second one with $n < 0$.

Another example of information that was missing is the set of live variables.
Indeed, in LLVM, live variables computation is processed during the register
allocation. Then, the data structure we work on doesn't have live variables
informations yet. To solve this problem, we coded our own pass, that is aimed at
computing these live variables after the transformation in SSA form.

@@

	Additionally, our analyser has no memory model:
\begin{itemize}
\item \emph{Store} instructions, that store a value in the memory, has no effect
in the analysis.
\item \emph{Load} instructions, that get a value $v$ in the memory and assign it
to a variable $x$, doesn't provide properties about the value $v$. Then, our
analyser considers $x$ can take all values of its type.
\end{itemize}
In the case of local variables, the transformation of the internal
representation into SSA form removes the \emph{Load/Store} instructions.
The use of global variables instead of local ones will induce huge
loss of precision, since there will still be \emph{Load/Store} instructions in
the IA.

@@

We tried our technique on example with a huge number of paths: ex8, ex8b and
ex8c (see Table \ref{npaths}). These benchmarks are constructed such that the
loop contains only a few
possible paths, but a lot of unfeasible paths. The idea is to show that our
technique doesn't blow up, even with a huge expanded multigraph. Results show
that in such benchmarks, our technique finds a much better invariant and is even
faster than \emph{Lookahead Widening}.

>>>>

Static analysis aims at automatically computing properties on programs, such as
the possible values of their variables during execution. This allows to show for
instance that a program will not overflow, will not divide by zero, and to compute
loop invariants\dots
The main applications of static analysis are compile-time optimisations, and
the proof of safety properties in critical systems, such as avionics.

@@

In this report we make the following contributions:
\begin{itemize}
\item We present a new technique, refered as \emph{Path Focusing}, that takes
benefits of the recent advances in the field of SMT-solving to improve precision
of static analysis by abstract interpretation. We also present an adaptation of
this technique to compute disjunctive invariants.
\item We present an implementation of this technique into a small analyser, as
well as the implementation of another technique proposed in \cite{GopanR06},
such that we can compare precision and cost of these two methods.
\end{itemize}

@@

Static analysis is aimed at discovering some properties about programs. In this
example, we would like to compute the set of possible values for the variable $x$ during execution.
The graph at the right side represents the program. The nodes of the graph are
called \emph{program points}. We can compute for instance the set $Y_1$ of
values for $x$ at point $p_1$. This set depends on the sets $Y_0$ and $Y_2$,
since there are two edges arriving at $p_1$: the first one comes from $p_0$, and
the second one from $p_2$.

@@

However, most of the time, the solution of this fixpoint equation must be
computed in a complex domain: in program analysis, this could be the state space
of the program, i.e the set of all possible states of the program. This
computation quickly becomes too costly or does not terminate.

@@

Termination of the fixpoint computation has to be guaranteed. This termination
depends on the properties of the abstract domain:
this one should be finite, of finite height,
or more generally satisfy the \emph{ascending chain condition},
meaning that there does not exist
any sequence $(x_i)_{i \geq 0}$ of elements of the abstract domain $A$ such that
$\forall i \geq 0,\ x_i <_A x_{i+1}$. Indeed, otherwise, the least fixpoint
computation could run indefinitely, since the ascending sequence of elements of
$A$ is infinite, and the fixpoint in never reached.

Many of the domains that are used for program analysis, including those of
intervals and convex polyhedra, do not satisfy this ascending chain condition. To ensure convergence in this case, another approximation
is performed: a new operator is defined, called \emph{widening operator}, that
extrapolates the limit of a sequence of abstract values
\cite{CC77,CousotCousot92-4}. This \emph{widening
operator} is usually noted $\widening: A \times A \rightarrow A$, and satisfies
the following properties:

@@

Let $x_1, x_2, \dots ,x_n$ be the numerical variables of a program (assume they
 are all in $\Q$). A state of
the program is then a point $\overrightarrow{x} \in \Q^n$.

@@

There have been various
work proposing a definition or a refinement of the widening operator in LRA
\cite{CH78,Hal79,HPR97,BlanchetCousotEtAl_PLDI03}, in order to fight
against loss of precision:
\begin{itemize}
\item \emph{Delayed widening}: instead of applying widening at the first
iteration of the analysis, one apply it after a number $n$ of iterations. During
the $n-1$ first iterations, only a convex hull is computed. In some cases,
delaying the widening operation is more precise. The algorithm still terminates,
since widening is applied within a finite time after $n$ iterations. In most of
the cases, $n$ can be chosen equal to $2$.
\item \emph{Widening with threshold (or ``up to'')}: In some cases, instead of
brutally applying the widening operator, a fixed set of linear inequalities $T$
can be chosen, and the widening operator $\widening_T$ is defined as follows:
$P \widening_T Q$ is the intersection of $P \widening Q$ with all the inequalities in
$T$ satisfied by both $P$ and $Q$.

@@

To each state $p \in P$ of the control flow graph, we associate an abstract
value $X_p \in D$, $D$ being in our case the domain of convex polyhedra over
$\Q^n$. Since the exact operation $\tau$ may not be expressible in the abstract
domain, we abstract it into $\tau^\#$ such that $\forall X \in D, \tau(X)
\subseteq \tau^\#(X)$.

@@

After a finite number of steps, the computation has reached the fixpoint and the
value $X_p$ at the end gives various linear relations between the
different numerical variables of the program at the control point $p$.

\subsection{Precision of the analysis}
In order to guarantee the termination of the computation, there is no need to
apply the widening operator at each control point $p \in P$. It is sufficient to
apply it on a subset of $P$, called $P_w$, such that removing the nodes of $P_w$
disconnects every cycles. For instance, $P_w$ can be chosen as the set of the
loop headers.

\subsubsection{Iteration strategies}
There exist several iteration strategies for computing the fixpoint: whatever
order we choose for updating the different $X_p, p\in P$, the result at the
end of the analysis is correct. Yet, the precision of the result and the time
before convergence can be very different. 
There have been some work to find efficient iteration strategies
\cite{Bou92}, such as first stabilising innermost loops, or stabilising strongly
connected components of the graph.

@@

	If the problem has a solution,  the SMT-solver will be in position to show
	which path in the control flow graph is selected, by giving a model: an
	assignments of the Boolean variables so that the formula is true.
	Among these Boolean variables, some of them are associated to a transition
	in the control flow graph:
	\begin{itemize}
	\item if the model has set to \emph{true} the Boolean variable $b_e$ associated
	to the transition $e$, then the focus path goes through this transition.
	\item in the other case, the focus path does not go through this transition.
	\end{itemize}

@@

\begin{itemize}
\item Step 0: Initially, $X_0 = \top, X_1 = \perp, X_7 = \perp$, and $A =
\{p_0\}$.
\item Step 1: We start in $p_0$. Is there a path starting in $p_0$, that arrives
in a state in $P_R$ and makes its abstract value grow? The SMT-solver answers
\emph{yes}, and gives the path $p_0 \rightarrow p_1$. We update $X_1$:
it is the polyhedron $x=y=0$. We update $A = \{p_1\}$. 
We do another SMT query: is there another path
starting in $p_0$ that makes an abstract value grow? The answer is no.
\item Step 2: We now work on $p_1$. Is there a path starting in $p_1$ that
arrives in a successor of $p_1$, and makes its abstract value grow? Answer is
yes, and the model given by the SMT-solver is the path $p_1 \rightarrow p_2
\rightarrow p_4 \rightarrow p_5 \rightarrow p_6 \rightarrow p_1$. Actually, this
is the first phase of the loop. 
The image of $x=y=0$ by this transition is $x=y=1$. The new $X_1$ is then $x=y
\wedge 0\leq x \leq 1$. The path is a self loop, but is seen for the first time,
so we do not apply widening yet. So, we redo an SMT-query, that gives us exactly
the same path. Then, we can do a widening/narrowing sequence. 
After widening, we have $x=y \wedge x \geq 0$, and narrowing gives $X_1 = (x=y)
\wedge (0 \leq x \leq 51)$.
\item Step 3: We still are in $p_1$: we ask for another focus path, and the
SMT-solver finds the path $p_1 \rightarrow p_3
\rightarrow p_4 \rightarrow p_5 \rightarrow p_6 \rightarrow p_1$. This is the
second phase of the loop, which is possible since $(x,y)=(51,51) \in X_1$.
Again, this is a self-loop, but we do not widen yet. The
image of $X_1$ by the transition is the single point $(52,50)$. $X_1$ is then
the convex hull of this point with the old $X_1$. A new SMT query gives the same
path, and we can now apply widening/narrowing. We obtain
$X_1 = (x \leq y) \wedge (102 - x - y \geq 0)$ after widening, and narrowing
gives $X_1 = (x \leq y) \wedge (102 - x - y \geq 0) \wedge y \geq 0$.
We can see that $X_1$ is the smallest polyhedron containing the possible values
for $(x,y)$ (see Figure \ref{result}.b).
\item Step 4: Again, we ask the SMT-solver for a new focus path, and it returns
$p_1 \rightarrow p_3 \rightarrow p_4 \rightarrow p_7$. The image of the
polyhedra $X_1$ after this path transformation is the single point $y=-1 \wedge
x = 102$. We can update $X_7 = {(102,-1)}$. Path focusing technique gives us the
precise result of this program. After that, the SMT-solver cannot find any new
path, so the analysis terminates.
\end{itemize}

@@

The new algorithm we propose is described in Algorithm \ref{gulwani2}.
A difference is that here, we do not know the number of paths. Then, the mapping
function $\sigma$ has to be defined on $\{1,..,m\} \times \N$. We could also
easily compute this number of paths when computing the $\rho$ formula.

@@

\paragraph{Path Focusing} 
The analysis of the loop starts with 
$p_1: x=i=r=0$. Then, the SMT-solver discovers the path $p_1 \rightarrow p_2
\rightarrow p_3 \rightarrow p_5 \rightarrow p_7 \rightarrow p_1$. We do a
widening/narrowing sequence and we find 
$ 0 \leq i \leq 50 \wedge r=0 \wedge -1 \leq x \leq 0$ at $p_1$. The SMT-solver
gives another path: 
$p_1 \rightarrow p_2
\rightarrow p_4 \rightarrow p_5 \rightarrow p_7 \rightarrow p_1$. After
analysing this path, we obtain 
$ 0 \leq i \leq 100 \wedge r=0 \wedge x=0$. Analysis terminates since the
SMT-solver gives $unsat$ answers. Here, We never focused on paths going through
point $p_6$, so our analysis gives us the invariant $x=0$ at the end of the
program. Indeed, there is no feasible paths going through $p_6$, but
\emph{Lookahead Widening} do not know it because of the convex hull at point
$p_5$.

@@

LLVM internal representation can be applied transformation passes that modifies
parts of it. These passes are mostly used for optimisation purpose. In our
analyser, we apply some already existing passes before running our 
abstract interpretation pass. Among the passes we run, we could cite:
\begin{itemize}
\item the \emph{PromoteMemoryToRegister} Pass, that transforms the internal
representation by promoting the memory variables to Static Single Assignment
(SSA) registers. This pass transform the control flow graph such that all the
variables are in SSA form, and removes most of the \emph{Load/Store} operations
into memory. It is useful since our analyser does not have a memory model, and
then loses lots of precision in the case of \emph{Load/Store} instructions.
\item the \emph{LoopSimplify} and \emph{LoopInfo} Passes, that transform all loops into a canonical
format, and compute some information related to loops (for instance, the set of
BasicBlocks that are parts of the loop, etc.).
\item the \emph{LowerSwitch} Pass, that transform a switch instruction into its
corresponding \emph{if-then-else} instructions. After running this pass, there
is no more \emph{Switch} instruction in the control flow graph, so the
implementation of the static analysis is easier.
\end{itemize}

@@

This is the case for the distinction between \emph{signed} and \emph{unsigned}
integers, which is nonexistent in the representation, since the compiler does
not
need this information to generate code. Yet, for static analysis, 
this distinction is required to improve precision:
for instance, if $n$ is unsigned, the constraint $n \geq 0$ could be added in
	our abstract value at the beginning of the analysis.
	This lack of precision may be limited by using disjunctive invariants: one
	could attach two polyhedra to the control point: the first one with the
	constraint $n \geq 0$, and the second one with $n < 0$.

Another example of information that was missing is the set of live variables.
Indeed, in LLVM, live variables computation is processed during the register
allocation. Then, the data structure we work on does not have live variables
informations yet. To solve this problem, we coded our own pass, that is aimed at
computing these live variables after the transformation in SSA form.

@@

	Additionally, our analyser has no memory model:
\begin{itemize}
\item \emph{Store} instructions, that store a value in the memory, has no effect
in the analysis.
\item \emph{Load} instructions, that get a value $v$ in the memory and assign it
to a variable $x$, does not provide properties about the value $v$. Then, our
analyser considers $x$ can take all values of its type.
\end{itemize}
In the case of local variables, the transformation of the internal
representation into SSA form removes the \emph{Load/Store} instructions.
The use of global variables instead of local ones will induce huge
loss of precision, since there will still be \emph{Load/Store} instructions in
the IA.

@@

We tried our technique on example with a huge number of paths: ex8, ex8b and
ex8c (see Table \ref{npaths}). These benchmarks are constructed such that the
loop contains only a few
possible paths, but a lot of unfeasible paths. The idea is to show that our
technique does not blow up, even with a huge expanded multigraph. Results show
that in such benchmarks, our technique finds a much better invariant and is even
faster than \emph{Lookahead Widening}.
